---
title: "Modern Applyed Statistics(Chap 11)"
output: pdf_document
---

```{r}
library(MASS)
library(class)
library(fastICA)
library(cluster)
options(width=65, digits=5)

# install.packages("../package/xgobi_1.2-15.tar.gz", repos = NULL, type = "source")
# install.packages("../package/RGtk2_2.20.36.tar.gz", repos = NULL, type = "source")
# install.packages("../package/rggobi_2.1.22.tar.gz", repos = NULL, type = "source")
```


## 11.1  Visualization methods

### 1) Principal Component analysis
```{r, fig.height=3}
# Iris data
ir <- rbind(iris3[,,1], iris3[,,2], iris3[,,3])
ir.species <- factor(c(rep("s", 50), rep("c", 50), rep("v", 50)))

# Principal Component for the log-transforrned iris data.
(ir.pca <- princomp(log(ir), cor = TRUE))
summary(ir.pca)
plot(ir.pca)
```

```{r, fig.height=4}
# First two principal components for the log-transforrned iris data.
ir.pc <- predict(ir.pca)
eqscplot(ir.pc[, 1:2], type = "n",
         xlab = "first principal component",
         ylab = "second principal component")
text(ir.pc[, 1:2], labels = as.character(ir.species),
     col = 3 + unclass(ir.species))
```

```{r}
# Crabs data
lcrabs <- log(crabs[, 4:8])
crabs.grp <- factor(c("B", "b", "O", "o")[rep(1:4, each = 50)])

# Principal Component for the crabs data.
(lcrabs.pca <- princomp(lcrabs))
loadings(lcrabs.pca)
lcrabs.pc <- predict(lcrabs.pca)
dimnames(lcrabs.pc) <- list(NULL, paste("PC", 1:5, sep = ""))
```

```{r, fig.height=4}
# First two principal components for the crabs data.
eqscplot(lcrabs.pc[, 1:2], type = "n",
         xlab = "first principal component",
         ylab = "second principal component")
text(lcrabs.pc[, 1:2], labels = as.character(crabs.grp),
     col = 3 + as.integer(crabs.grp)) 
```


### 2) Exploratory projection pursuit
```{r}
if(FALSE) { ## needs interaction with XGobi, or, better, rggobi
  library(xgobi)
  xgobi(lcrabs, colors = c("SkyBlue", "SlateBlue", "Orange",
                           "Red")[rep(1:4, each = 50)])
  xgobi(lcrabs, glyphs = 12 + 5*rep(0:3, each = 50, 4))
  
  library(rggobi)
  g <- ggobi(lcrabs)
  d <- displays(g)[[1]]
  pmode(d) <- "2D Tour"
  crabs.grp <- factor(c("B", "b", "O", "o")[rep(1:4, each = 50)])
  glyph_colour(g$lcrabs) <- crabs.grp
  colorscheme(g) <- "Paired 4"
}
```  

\newpage

### 3) Distance methods
```{r}
# Distance-based representations of the iris data
ir.scal <- cmdscale(dist(ir) , k = 2, eig = T)

distp <- dist(ir); dist2 <- dist(ir.scal$points)
sum((distp - dist2)^2)/sum(distp^2) # calculating a measure of 'stress'
```


```{r, fig.height=4, fig.width=5, message=FALSE, results='hide'}
par(mfrow = c(2,2))

eqscplot(ir.scal$points, type = "n", main = "Metric scaling", ratio = 2)
text(ir.scal$points, labels = as.character(ir.species), 
     col = 3 + as.integer(ir.species), cex = 0.8)
ir.sam <- sammon(dist(ir[-143,]))
eqscplot(ir.sam$points, type = "n", main = "Sammon mapping", ratio = 2)
text(ir.sam$points, labels = as.character(ir.species[-143]), 
     col = 3 + as.integer(ir.species), cex = 0.8)
ir.iso <- isoMDS(dist(ir[-143,]))
eqscplot(ir.iso$points, type = "n", main = "Kruskal's MDS", ratio = 2)
text(ir.iso$points, labels = as.character(ir.species[-143]), 
     col = 3 + as.integer(ir.species), cex = 0.8) 
```

\newpage

```{r, fig.height=4, fig.width=6, message=FALSE, results='hide'}
# Sammon mapping of crabs data
cr.scale <- 0.5 * log(crabs$CL * crabs$CW)
slcrabs <- lcrabs - cr.scale
cr.means <- matrix(0, 2, 5)
cr.means[1,] <- colMeans(slcrabs[crabs$sex == "F", ])
cr.means[2,] <- colMeans(slcrabs [crabs$sex == "M", ])
dslcrabs <- slcrabs - cr.means[as.numeric(crabs$sex),]
lcrabs.sam <- sammon(dist(dslcrabs))
eqscplot(-lcrabs.sam$points, type = "n", xlab = "", ylab = "", tol = 0.08, ratio = 1.5)
text(-lcrabs.sam$points , labels = as.character(crabs.grp), 
     col = rep(c("blue", "orange"), each = 100))
```


```{r, fig.height=4, results=FALSE}
# Isotonic multidimensional scaling representation of the fgl data.
fgl.iso <- isoMDS(dist(as.matrix(fgl[-40, -10])))
eqscplot(fgl.iso$points, type = "n", xlab = "", ylab = "", axes = FALSE)
# either
# for(i in seq(along = levels(fgl$type))) {
#   set <- fgl$type[-40] == levels(fgl$type)[i]
#   points(fgl.iso$points[set,], pch = 18, cex = 0.6, col = 2 + i)}
# key(text = list(levels(fgl$type), col = 3:8))
# or
text(fgl.iso$points, labels = c("F", "N", "V", "C", "T", "H")[fgl$type[-40]], cex = 0.6)
```


```{r, results=FALSE}
fgl.iso3 <- isoMDS(dist(as.matrix(fgl[-40, -10])), k = 3)
# S: brush(fgl.iso3$points)
fgl.col <- c("SkyBlue", "SlateBlue", "Orange", "Orchid", "Green", "HotPink")[fgl$type]
# xgobi(fgl.iso3$points, colors = fgl.col)
```
\newpage


### 4) Self-organizing maps
```{r}
# Batch SOM applied to the crabs dataset.
set.seed(0)
gr <- somgrid(topo = "hexagonal")
crabs.som <- batchSOM(lcrabs, gr, c(4, 4, 2, 2, 1, 1, 1, 0, 0))

# stars plot of the representatives
stars(crabs.som$codes, labels = NULL, frame.plot = T)
```


```{r, fig.height=4.5}
set.seed(0)

# Plot that shows the assignments of the original points
bins <- as.numeric(knn1(crabs.som$code, lcrabs, 0:47))
plot(crabs.som$grid, type = "n", frame.plot = T,
     xlim = c(min(crabs.som$grid$pts[,1])-0.4, max(crabs.som$grid$pts[,1])+0.4),
     ylim = c(min(crabs.som$grid$pts[,2])-0.4, max(crabs.som$grid$pts[,2])+0.4))
symbols(crabs.som$grid$pts[, 1], crabs.som$grid$pts[, 2], 
        circles = rep(0.4, 48), inches = FALSE, add = TRUE)
text(crabs.som$grid$pts[bins, ] + rnorm(400, 0, 0.1), as.character(crabs.grp))
```

```{r, fig.height=3.5}
set.seed(0)

# Traditional SOM applied to the crabs dataset.
crabs.som2 <- SOM(lcrabs, gr); stars(crabs.som2$codes, frame.plot = T)
```

\newpage

### 5) Biplots
```{r, fig.height=7, fig.width=7}
# Principal component biplot of the part of the state.x77 data.
state <- state.x77[, 2:7]; row.names(state) <- state.abb
state.pca <- princomp(state, cor = TRUE)
state.pca$loadings[,2] <- -state.pca$loadings[,2]
state.pca$scores[,2] <- -state.pca$scores[,2]
biplot(state.pca, pc.biplot = TRUE, cex = 0.7, expand = 0.8)
```

\newpage

### 6) Independent component analysis
```{r, fig.width=8, fig.height=4}
set.seed(0)

# Boxplots of four ’signals’ recovered by ICA from the crabs data.
nICA <- 4
crabs.ica <- fastICA(crabs[, 4:8], nICA)
Z <- crabs.ica$S
par(mfrow = c(1, nICA))
for(i in 1:nICA) boxplot(Z[, i] ~ crabs.grp)
```

\newpage

### 7) Glyph representations
```{r, fig.width=8, fig.height=8}
# stars plot of the state.x77 dataset.
stars(state.x77[, c(7, 4, 6, 2, 5, 3)])
```

\newpage

### 8) Parallel coordinate plots
```{r, fig.width=8}
# Parallel coordinates plots of the state.x77 dataset.
parcoord(state.x77[, c(7, 4, 6, 2, 5, 3)])

# Parallel coordinates plots of the log-transforrned iris data
parcoord(log(ir)[, c(3, 4, 2, 1)], col = 1 + (0:149)%/%50)
```

\newpage

## 11.2 Cluster Analysis
```{r, fig.width=8, fig.height=3.5}
# Dendograms for the socio-economic data on Swiss provinces by single-link clustering
swiss.x <- as.matrix(swiss[,-1])
h <- hclust(dist(swiss.x), method = "single")
plot(h, labels = h$order, main = "", xlab = "", sub = "")
```


```{r, fig.width=8, fig.height=3.5}
# Dendograms for the socio-economic data on Swiss provinces by divisive clustering
d <- diana(swiss.x, )
pltree(d, labels = d$order, main = "", xlab = "", sub = "")
```

\newpage

```{r, fig.height=6, fig.width=8, results=FALSE, message=FALSE}
# First two principal components for the swiss data 
# and labeling by the groups assigned by K-means
h <- hclust(dist(swiss.x), method = "average")
initial <- tapply(swiss.x, list(rep(cutree(h, 3), ncol(swiss.x)), col(swiss.x)), mean)
dimnames(initial) <- list(NULL, dimnames(swiss.x)[[2]])
km <- kmeans(swiss.x, initial)
(swiss.pca <- princomp(swiss.x))
swiss.px <- predict(swiss.pca); swiss.px[,2] <- -swiss.px[,2] 
dimnames(km$centers)[[2]] <- dimnames(swiss.x)[[2]]
swiss.centers <- predict(swiss.pca, km$centers); swiss.centers[,2] <- -swiss.centers[,2]
eqscplot(swiss.px[, 1:2], type = "n", 
         xlab = "first principal component" , ylab = "second principal component")
text(swiss.px[, 1:2], labels = km$cluster)
points(swiss.centers[,1:2], pch = 3, cex = 5)
identify(swiss.px[, 1:2], cex = 0.5)
```


\newpage



```{r, fig.height=8, fig.width=12, results=FALSE, message=FALSE}
# Clusterings of the Swiss provinces data by pam, me, emclust
par(mfrow = c(2,2))

swiss.pam <- pam(swiss.px, 3)
eqscplot(swiss.px[, 1:2], type = "n",
         xlab = "first principal component", ylab = "second principal component",
         main = "pam")
text(swiss.px[,1:2], labels = swiss.pam$clustering)
points(swiss.pam$medoid[,1:2], pch = 3, cex = 3)

library(mclust)
vals <- mclustBIC(swiss.x)
sm <- summary(vals, swiss.x)
eqscplot (swiss.px [, 1: 2], type = "n",
          xlab = "first principal component" , ylab = "second principal component", 
          main = "emclust")
text(swiss.px[, 1:2], labels = sm$classification) 

h <- hc(modelName = "VVV", swiss.x)
mh <- as.vector(hclass(h, 3))
z <- me(modelName = "VVV", swiss.x, z = 0.5*(unmap(mh)+1/3))
eqscplot(swiss.px[, 1:2], type = "n", 
         xlab = "first principal component", ylab = "second principal component",
         main = "me")
text(swiss.px[, 1:2], labels = max.col(z$z)) 
```

\newpage

# 11.3 Factor analysis
```{r}
ability.FA <- factanal(covmat = ability.cov, factors = 1)
ability.FA
(ability.FA <- update(ability.FA, factors = 2))
#summary(ability.FA)
round(loadings(ability.FA) %*% t(loadings(ability.FA)) +
        diag(ability.FA$uniq), 3)
```


```{r, fig.width=6, fig.height=6}
# Factor rotations
library(GPArotation)
L <- loadings(ability.FA)
print(oblirot <- oblimin(L))
par(pty = "s")
eqscplot(L, xlim = c(0,1), ylim = c(0,1))
if(interactive()) identify(L[1:6,1], dimnames(L)[[1]])
naxes <- oblirot$Th
arrows(rep(0, 2), rep(0, 2), naxes[,1], naxes[,2])
text(L[1:6,1:2], dimnames(L)[[1]])
```


# 11.4 Discrete multivariate analysis
## mosaic plot
```{r, fig.height=12, fig.width=6}
par(mfrow = c(2,1))
# Mosaic plots for Fisher's data on people from Caithness
caith <- as.matrix(caith)
names(dimnames(caith)) <- c("eyes", "hair")
mosaicplot(caith, color = TRUE)
# Mosaic plots for Copenhagen housing satisfaction data
House <- xtabs(Freq ~ Type + Infl + Cont + Sat, housing)
mosaicplot(House, color = TRUE)
```


```{r}
corresp(caith)
```

\newpage

```{r, fig.width=12, fig.height=4}
# Three variants of correspondence analysis plots from Fisher's data
caith2 <- caith
dimnames(caith2)[[2]] <- c("F", "R", "M", "D", "B")
par(mfcol = c(1, 3))
plot(corresp(caith2, nf = 2)); title("symmetric")
plot(corresp(caith2, nf = 2), type = "rows"); title("rows")
plot(corresp(caith2, nf = 2), type = "col"); title("columns")
```


```{r}
# Multiple correspondence analysis plot of dataset farms
farms.mca <- mca(farms, abbrev = TRUE)  # Use levels as names
plot(farms.mca, cex = rep(0.7, 2))
```

